---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 8,
    fig.height = 6
)
```

```{r setup}

library(sentinelBurnR)
library(terra)

```

# Getting started with sentinelBurnR

`sentinelBurnR` provides tools for analysing wildfire and its
environmental context using satellite and climate data.

The package uses Sentinel-2 multispectral imagery to identify burned
areas and examine vegetation condition. It can also compare vegetation
moisture with historical observations and place these observations in
the context of climate data such as rainfall.

A typical analysis can include:

1. defining an area of interest;
2. identifying suitable Sentinel-2 imagery;
3. comparing pre- and post-fire imagery;
4. analysing vegetation condition;
5. assessing vegetation moisture anomalies;
6. examining recent rainfall and other climate conditions.

This vignette uses data from the Dunwich area of Suffolk, UK, associated
with a wildfire in summer 2026. All example data are included with the
package, so the vignette can be run without downloading data.

# True-colour imagery

Sentinel-2 records reflected light in several parts of the
electromagnetic spectrum. The simplest way to view the imagery is as a
true-colour image, combining the red, green and blue wavelengths to
produce an image similar to that seen by the human eye.

The following images show the Dunwich area before and after the fire.

```{r true-colour-pre, fig.width=10, fig.height=8}

pre_visual <- demo_data("visual_pre")

terra::plotRGB(
    pre_visual,
    r = 1,
    g = 2,
    b = 3,
    stretch = "lin"
)
```

```{r true-colour-post, fig.width=10, fig.height=8}
post_visual <- demo_data("visual_post")

terra::plotRGB(
    post_visual,
    r = 1,
    g = 2,
    b = 3,
    stretch = "lin"
)
```

The effects of the fire can be seen in the true-colour imagery, but
visible colour alone is not an ideal way to determine the extent or
severity of burning. Soil, shadows, vegetation type and naturally bare
ground can produce similar changes in appearance.

Sentinel-2 also measures near-infrared and short-wave infrared light,
which provide much stronger information about changes in vegetation
following fire. These wavelengths are used to calculate the Normalized
Burn Ratio (NBR).

Sentinel-2 also measures wavelengths beyond those visible to the human
eye. The Normalized Burn Ratio (NBR) uses near-infrared and short-wave
infrared measurements, which respond strongly to changes in vegetation
and surface condition following fire.

# Demo data

Load the area of interest and the Sentinel-2 collections:

```{r demo-data}
demo_dir <- system.file(
        "extdata",
        "demo",
        package = "sentinelBurnR"
    )
aoi <- readRDS(
            file.path(demo_dir, "demo_aoi.rds")
        )

pre <- demo_data("pre")
post <- demo_data("post")
```

The collections contain cropped Sentinel-2 imagery for the Dunwich
area. In a normal analysis these collections would instead be created
using `search_s2()` and `download_s2()`.

For example, the corresponding workflow is:

```{r sentinel-workflow, eval=FALSE}
pre_search <- search_s2(
    aoi,
    start = "2026-07-14",
    end = "2026-07-29"
)

post_search <- search_s2(
    aoi,
    start = "2026-07-29",
    end = "2026-08-22"
)

pre <- download_s2(
    pre_search,
    assets = s2_burn_assets
)

post <- download_s2(
    post_search,
    assets = s2_burn_assets
)
```

The code is not evaluated in this vignette because searching and
downloading Sentinel-2 imagery requires an internet connection.

# Burn analysis

Burn severity can be estimated from changes in the Normalized Burn
Ratio (NBR). NBR contrasts near-infrared and short-wave infrared
reflectance. Healthy vegetation generally has high near-infrared
reflectance, whereas recently burned surfaces tend to have relatively
higher short-wave infrared reflectance.

`analyse_burn()` calculates pre- and post-fire NBR and their difference
(dNBR):

```{r burn-analysis}
burn <- analyse_burn(
    pre = pre,
    post = post,
    boundary = aoi
)

burn
```

The resulting `sbr_burn` object contains the pre- and post-fire
composites, NBR rasters, dNBR, the detected burned area, burn severity
and processing provenance.

The spatial pattern of dNBR can be displayed with:

```{r plot-dnbr}

plot_index(
    burn$dnbr,
    index = "dnbr",
    boundary = aoi$geometry
)
```

Burn severity can also be mapped directly:

```{r plot-severity}

plot_severity(
    burn$severity,
    boundary = aoi$geometry
)
```

# Vegetation condition

The same pre-fire Sentinel-2 imagery can be used to examine vegetation
condition.

`analyse_vegetation()` calculates three complementary spectral indices:

- **NDVI** — an indicator of vegetation greenness;
- **NDMI** — sensitive to vegetation moisture;
- **MSI** — a moisture-stress index based on near- and short-wave
  infrared reflectance.

```{r vegetation-analysis}
vegetation <- analyse_vegetation(
    pre
)

vegetation
```

For example, pre-fire vegetation moisture can be mapped using NDMI:

```{r vegetation-ndmi}
plot_index(
    vegetation$ndmi,
    index = "ndmi"
)
```

These indices describe vegetation condition at the time of observation.
A low NDMI value alone does not demonstrate drought: vegetation moisture
varies naturally between vegetation types and through the growing
season. To assess whether conditions were unusually dry, the current
observation can instead be compared with observations from previous
years.

# Vegetation moisture anomaly

Vegetation moisture at a single point in time is difficult to interpret
in isolation. NDMI varies between vegetation types and also changes
seasonally.

To provide historical context, `sentinelBurnR` can compare current NDMI
with observations from the same part of the growing season in previous
years.

For this example, the package includes annual NDMI composites for
2018--2025 and a pre-fire observation from 29 July 2026.

```{r drought-data}
historical_ndmi <- demo_data("drought_historical")
current_ndmi <- demo_data("drought_current")
```

The historical data contain one seasonal NDMI composite for each year:

```{r drought-years}
names(historical_ndmi)
```

For the demonstration we can analyse these already-prepared rasters:

```{r drought-analysis}
drought <- sentinelBurnR:::analyse_drought_rasters(
    annual = historical_ndmi,
    current = current_ndmi[["2026-07-29"]],
    current_date = "2026-07-29"
)

drought
```

The anomaly is calculated as the difference between current NDMI and
the historical seasonal baseline. Negative values therefore indicate
vegetation that was less moisture-rich than expected for the time of
year.

For the Dunwich example, median NDMI on 29 July 2026 was approximately
0.05 below the 2018--2025 seasonal baseline. This indicates relatively
dry vegetation conditions before the fire, although the anomaly should
not by itself be interpreted as a direct measure of drought severity.

```{r drought-map}
plot_drought(
    drought,
    index = "anomaly",
    boundary = aoi$geometry
)
```

A standardised anomaly is also available:

```{r drought-standardised}
plot_drought(
    drought,
    index = "standardised",
    boundary = aoi$geometry
)
```

The standardised anomaly expresses the departure relative to historical
interannual variability. Where historical variability is very small,
standardisation can become unstable, so `sentinelBurnR` masks pixels
whose historical standard deviation falls below a minimum threshold.
Absolute NDMI anomaly therefore remains an important part of the
interpretation.

# Rainfall and climate context

The Sentinel-2 moisture anomaly indicates whether vegetation was
unusually dry compared with previous years. It does not, by itself,
tell us why the vegetation was dry.

Meteorological observations provide an independent line of evidence.
`sentinelBurnR` can compare recent rainfall and temperature with the
equivalent periods during a historical climate baseline. Here we use
1991--2020 as the reference period.

For a full analysis, `get_climate_baseline()` downloads the required
ERA5 data for both the analysis year and the baseline years. Only the
months needed to calculate the requested climate windows are retrieved.

```{r climate-download, eval=FALSE}
climate_data <- sentinelBurnR::get_climate_baseline(
    boundary = aoi,
    date = as.Date("2026-07-29"),
    baseline_years = 1991:2020
)

climate <- sentinelBurnR::analyse_climate(
    rainfall = climate_data$rainfall,
    temperature = climate_data$temperature,
    date = as.Date("2026-07-29"),
    baseline_years = 1991:2020,
    windows = c(30, 60, 90)
)

climate
```

Downloading ERA5 data requires access to the Copernicus Climate Data
Store. To keep this vignette reproducible without network access, the
executable example below uses rainfall data bundled with the package.

```{r climate-analysis}
rainfall <- demo_data("rainfall")

climate <- sentinelBurnR::analyse_climate(
    rainfall = rainfall,
    date = as.Date("2026-07-29"),
    baseline_years = 1991:2020,
    windows = c(30, 60, 90)
)

climate$summary |>
    dplyr::transmute(
        `Period (days)` = window_days,
        `2026 rainfall (mm)` = current_mm,
        `1991–2020 median (mm)` = baseline_median_mm,
        `Rainfall anomaly (mm)` = anomaly_mm,
        `% of historical median` = percent_of_normal,
        `Historical percentile` = historical_percentile
    ) |>
    knitr::kable(
        digits = 1,
        caption = paste(
            "Rainfall before 29 July 2026 compared with",
            "the equivalent periods in the 1991–2020 baseline."
        )
    )
```

The percentage of normal rainfall compares the observed rainfall total
with the median of the equivalent periods in the baseline years. Values
below 100% indicate less rainfall than the historical median.

The historical percentile provides a complementary measure. It ranks
the current rainfall total against the equivalent periods in the 30
baseline years. With a 30-year baseline, the percentile is an empirical
ranking within a relatively small sample. A value of 0%, for example,
means that the current value was below all 30 baseline values; it should
not be interpreted as a literal estimate of a zero-percent
climatological probability.

The same comparison can be visualised directly.

```{r climate-rainfall-plot, fig.width=7, fig.height=5}
sentinelBurnR::plot_climate(
    climate,
    metric = "rainfall"
)
```

Rainfall substantially below the historical median provides
meteorological evidence that a negative vegetation moisture anomaly
occurred during an unusually dry period. The two measurements are
independent: Sentinel-2 describes the condition of the vegetation,
whereas ERA5 describes the meteorological conditions preceding the
satellite observation.

When temperature data are supplied to `analyse_climate()`, the same
analysis also compares recent temperature with the historical baseline.

```{r climate-temperature, eval=FALSE}
climate$temperature

sentinelBurnR::plot_climate(
    climate,
    metric = "temperature"
)
```

Temperature anomalies are expressed in degrees Celsius relative to the
baseline, while the historical percentile describes where the current
temperature lies among the equivalent periods in the baseline years.
As with rainfall, these percentiles are empirical rankings within the
baseline sample rather than estimates of climatological probability.

## Prolonged dry weather

Total rainfall does not tell us how rain was distributed through time.
A similar total could result from frequent small showers or from a
long period with almost no rain. `analyse_climate()` therefore also
examines the duration of dry spells.

```{r dry-spell}
climate$dry_spell
```

By default, a day with less than 1 mm of rainfall is treated as a dry
day. The result reports the number of dry days in the analysis window,
the longest consecutive dry spell, its start and end dates, and the
number of days since the most recent day with at least 1 mm of rain.

The current dry spell can also be compared with the equivalent period
in each baseline year.

```{r dry-spell-baseline}
climate$dry_spell_baseline[c(
    "median_days",
    "max_days",
    "percentile"
)]
```

The baseline median describes the typical longest dry spell during the
equivalent historical period. The percentile shows how the current
longest dry spell ranks among the baseline years.

Taken together, rainfall, temperature, dry-spell and satellite
observations provide complementary evidence about the environmental
conditions preceding a fire. Rainfall describes water input,
temperature influences evaporative demand, and Sentinel-2 measures the
condition of the vegetation itself.

These observations describe environmental conditions. They do not
establish the cause of a fire or, by themselves, quantify wildfire
risk.

## Landscape context

Wildfire behaviour is influenced not only by vegetation condition and recent
weather, but also by the spatial arrangement of fuels and other landscape
features. Interfaces between vegetation types can be particularly important,
as can roads, tracks and other features that may affect access, suppression
and fuel continuity.

`analyse_landscape()` provides tools for describing these relationships. It
does not calculate a wildfire risk score; instead, it reports measurable
landscape characteristics that can be interpreted alongside vegetation,
drought and climate information.

```{r landscape-data}
landcover <- demo_data("landcover")
habitat <- demo_data("habitat")
transport <- demo_data("transport")
```

For the Dunwich example, we examine the interface between lowland heathland
and coniferous land cover. Priority habitat polygons provide the heathland
mapping, while coniferous cover is identified from OS land-cover classes.

```{r landscape-classes}
heath <- habitat |>
    dplyr::filter(
        stringr::str_detect(
            mainhabs,
            "Lowland heathland"
        )
    )

conifer <- landcover |>
    dplyr::filter(
        purrr::map_lgl(
            oslandcovertierb,
            \(x) any(
                x %in% c(
                    "Coniferous Trees",
                    "Scattered Coniferous Trees"
                )
            )
        )
    )
```


```{r landscape-analysis}
landscape <- sentinelBurnR::analyse_landscape(
    landcover = landcover,
    category = "lcb",
    habitat = habitat,
    transport = transport,
    interfaces = list(
        heath_conifer = list(
            source = heath,
            target = conifer,
            source_label = "Lowland heathland",
            target_label = "Coniferous land cover"
        )
    ),
    distances = c(0, 10, 25, 50, 100)
)

landscape
```


The interface results are complementary measures. `interface_summary`
reports the cumulative length of heathland boundary with coniferous cover
within each specified distance. `interface_bands` reports the area of
coniferous cover in mutually exclusive distance bands around the heathland.

In this example, mapped coniferous cover is strongly concentrated close to
the heathland boundary. This describes the spatial relationship between the
two mapped cover types; it should not by itself be interpreted as a measure
of wildfire probability or expected fire behaviour.



```{r landscape-bands}
landscape$interface_bands |>
    dplyr::transmute(
        Interface = interface,
        `Distance from heath (m)` =
            paste0(lower_m, "–", upper_m),
        `Conifer area (ha)` = area_ha
    ) |>
    knitr::kable(
        digits = 2,
        caption =
            "Coniferous land cover by distance from mapped lowland heathland."
    )
```

o be compared with the vegetation interface.
In this demonstration area, very little of the heathland boundary near
conifer coincides with mapped transport features. A mapped road or track
should not automatically be interpreted as an effective firebreak: its
importance depends on characteristics such as width, surface, surrounding
vegetation, condition and accessibility.

```{r landscape-map, fig.width=9, fig.height=7, out.width="100%"}

aoi_sf <- aoi |>
    sentinelBurnR:::aoi_to_spatvector() |>
    sf::st_as_sf()

ggplot2::ggplot() +
    ggplot2::geom_sf(
        data = landcover,
        fill = "grey95",
        colour = NA
    ) +
    ggplot2::geom_sf(
        data = conifer,
        fill = "darkgreen",
        colour = NA,
        alpha = 0.7
    ) +
    ggplot2::geom_sf(
        data = heath,
        fill = "goldenrod",
        colour = "goldenrod4",
        alpha = 0.6
    ) +
    ggplot2::geom_sf(
        data = transport,
        fill = "grey30",
        colour = "grey20"
    ) +
    ggplot2::geom_sf(
        data = aoi_sf,
        fill = NA,
        colour = "black",
        linewidth = 0.5
    ) +
    ggplot2::coord_sf(datum = NA) +
    ggplot2::labs(
        title = "Landscape context",
        subtitle = "Lowland heathland, coniferous land cover and transport features",
        caption = "Dunwich demonstration area"
    ) +
    ggplot2::theme_minimal()
```










### NBR — Normalized Burn Ratio
NBR is a way of using satellite imagery to distinguish healthy vegetation from recently burned ground.
Healthy green vegetation reflects a lot of near-infrared light, while burned, dry and charred surfaces tend to reflect relatively more short-wave infrared light. NBR combines these two parts of the spectrum into a single number.
In general, high NBR indicates healthy vegetation, while lower NBR is associated with sparse, dry or burned vegetation.
The important point is that NBR alone does not tell us that a fire has occurred. Different vegetation and soil types naturally have different NBR values. It becomes much more useful when we compare images from before and after a fire.

### dNBR — change caused by the fire
dNBR measures how much NBR changed between the pre-fire and post-fire satellite images.
It is calculated approximately as:
\[
dNBR = NBR_{before} - NBR_{after}
\]If vegetation was healthy before the fire and subsequently burned, NBR falls and dNBR becomes positive.
So, broadly:
little change around zero → little evidence of burning;
larger positive values → a stronger change consistent with burning;
negative values → vegetation became greener or wetter rather than more burned.
This makes dNBR particularly useful for mapping the burn scar, because it measures change rather than simply looking for dark or bare ground.

### Burn severity
Burn severity turns the continuous dNBR values into categories that are easier to interpret on a map.
Pixels with a small change may be classified as unburned or very lightly affected, while increasingly large changes can be classified into progressively higher burn-severity classes.
The important distinction is that this is satellite-derived burn severity. It describes the magnitude of the change in vegetation and surface reflectance detected by Sentinel-2. It is not necessarily identical to ecological damage measured on the ground—for example, tree mortality, depth of soil heating or loss of particular species.
So a useful plain-English description is:
Burn severity shows where the fire produced the greatest change to the vegetation and ground surface visible from the satellite.

## #Vegetation condition
The vegetation analysis uses several satellite indices that respond to different aspects of vegetation condition.
NDVI mainly describes vegetation greenness and density. Healthy, actively growing vegetation generally has higher NDVI.
NDMI is more sensitive to vegetation moisture. Lower values can indicate drier vegetation, although naturally dry vegetation types can also have low NDMI.
MSI is another measure of vegetation moisture stress. Unlike NDMI, higher MSI generally corresponds to greater moisture stress.
These measurements are useful for asking questions such as “Was the vegetation unusually dry before the fire?” But a single satellite image cannot answer that very well, because heathland, woodland and grassland naturally have different values and all change through the seasons.

### Vegetation moisture anomaly
This is why the historical comparison is important.
A vegetation moisture anomaly asks whether the vegetation is wetter or drier than would normally be expected at that location and time of year.
For example, in the Dunwich analysis we compare the NDMI observed on 29 July 2026 with NDMI from the same part of the growing season in 2018–2025.
We calculate:
\[
\text{NDMI anomaly}
=
\text{current NDMI}
-
\text{historical NDMI}
\]Therefore:
zero → approximately normal;
positive → wetter/moister vegetation than usual;
negative → drier vegetation than usual.
At Dunwich the median anomaly was about −0.05, meaning vegetation moisture was generally lower than its recent historical seasonal baseline.

Standardised vegetation moisture anomaly
The absolute anomaly tells us how much NDMI changed. The standardised anomaly asks a slightly different question:
How unusual is that change compared with the amount this location normally varies from year to year?

It is essentially:
\[
\frac{\text{current NDMI} - \text{historical average}}
     {\text{historical variability}}
\]A value around −2, for example, means vegetation moisture is roughly two historical standard deviations below normal.
This can help distinguish a place that is slightly dry but naturally very variable from one experiencing conditions that are genuinely unusual for that location.
There is an important caveat: if a pixel normally varies very little, dividing by a very small historical variability can produce misleadingly extreme values. That's why sentinelBurnR masks locations where the historical standard deviation is very small.

### Drought anomaly versus drought
I'd make this distinction explicit in the vignette:
A negative vegetation moisture anomaly is evidence of unusually dry vegetation, but it is not by itself proof of meteorological drought.

Sentinel-2 tells us about the response of the vegetation. ERA5 rainfall tells us about the weather conditions that may have produced that response. In your Dunwich example, those two independent lines of evidence can then be considered together.
