---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

library(sentinelBurnR)
library(terra)

cache_dir <- file.path(
    path.expand("~"),
    "Library",
    "CloudStorage",
    "GoogleDrive-julian.flowers12@gmail.com",
    "My Drive",
    "Sentinel2",
    "cache"
)

```

# Introduction

`sentinelBurnR` provides tools for detecting and summarising burned areas from Sentinel-2 imagery, together with environmental context such as rainfall from ERA5.

Boundary
    │
    ▼
search_s2()
    │
    ▼
download_s2()
    │
    ▼
detect_burns()
    │
    ▼
burn_summary()
    │
    └──────────────► get_rainfall()




## Load data

```{r}
aoi <- read_aoi(
    system.file(
        "extdata",
        "dunwich.gpkg",
        package = "sentinelBurnR"
    )
)

boundary <- read_boundary(system.file(
        "extdata",
        "dunwich.gpkg",
        package = "sentinelBurnR"
    ))

rainfile <- system.file(
    "extdata",
    "era5",
    "2024_07.nc",
    package = "sentinelBurnR"
)

burn <- readRDS(
    system.file(
        "extdata",
        "burn",
        "burn.rds",
        package = "sentinelBurnR"
    )
)
```



## Search Sentinel-2 satellite imagery

```{r, eval = TRUE}

pre_images <- search_s2(
    aoi,
    start = "2026-07-14",
    end = "2026-07-29"
)

post_images <- search_s2(
    aoi,
    start = "2026-07-29",
    end = "2026-08-22"
)


```

## Download image

```{r, eval = TRUE}

## Download imagery

pre_collection <- download_s2(

    pre_images, 
    
    max_cloud = 20, 
    
    limit = 1,
    
    workers = 4

)

post_collection <- download_s2(

    post_images,
    
    max_cloud = 20,
    
    limit = 1,
  
    workers = 4

)
```

## Burn analysis

```{r}

burn <- analyse_burn(
    pre = pre_collection,
    post = post_collection,
    boundary = aoi
)


```

## Get rainfall

```{r, fig.width=9, fig.height=6}
r <- terra::rast(rainfile)

rain <- extract_rainfall(
    r,
    boundary = boundary
)

plot(rain)
```

Burn severity depends not only on vegetation but also on recent weather conditions. Rainfall in the days and weeks before a fire influences fuel moisture and can help explain patterns of burn severity observed in satellite imagery.

## Plot indices

```{r, fig.width=9, fig.height=6}

plot_index(
    burn$dnbr,
    index = "dnbr",
    boundary = boundary,
    caption = burn_caption(burn)
)

```

