---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

library(sentinelBurnR)
library(terra)
library(patchwork)

cache_dir <- file.path(
    path.expand("~"),
    "Library",
    "CloudStorage",
    "GoogleDrive-julian.flowers12@gmail.com",
    "My Drive",
    "Sentinel2",
    "cache"
)

```

# Introduction

`sentinelBurnR` provides an end-to-end workflow for producing burn severity maps from Sentinel-2 imagery.

This vignette demonstrates the complete workflow:

1. Create and read an Area of Interest (AOI)
2. Search the Sentinel-2 catalogue
3. Download imagery
4. Build cloud-free composites
5. Calculate NBR
6. Calculate dNBR

## Create an AOI

```{r}

lat <- 52.2522
lon <- 1.6271
  
buffer <- 2500

```


```{r}
aoi <- create_aoi(lat = lat, lon = lon, radius = buffer)
```


## Read an AOI

```{r}


aoi <- read_aoi(aoi$geometry)

```

## Search Sentinel-2

```{r}

pre_images <- search_s2(
    aoi,
    start = "2026-07-14",
    end = "2026-07-29"
)

post_images <- search_s2(
    aoi,
    start = "2026-07-29",
    end = "2026-08-16"
)


```

## Download

```{r, eval = TRUE}

## Download imagery

pre_collection <- download_s2(

    pre_images, limit = 10, max_cloud = 30,
    
    workers = 6

)

post_collection <- download_s2(

    post_images, limit = 20,
    
    max_cloud = 40,
  
    workers = 6

)
```

## Composite

```{r, eval=FALSE}

pre_comp <- build_composite(
    assets = c("red", "nir08", "swir22", "green", "blue"),
    pre_collection
)

post_comp <- build_composite(
    assets = c("red", "nir08", "swir22", "green", "blue"),
    post_collection
)

```

## NBR

```{r}

pre_nbr <- calc_nbr(
    pre_comp
)

post_nbr <- calc_nbr(
    post_comp

)

```

## dNBR

```{r}

dnbr <- calc_dnbr(

    pre_nbr,

    post_nbr
  

)

plot(dnbr)


```

## Classify

```{r}

severity <- classify_dnbr(dnbr) 

severity|>
  plot()

```



```{r}

p1 <- plot_rgb(
    pre_comp,
    title = "Pre-fire RGB",
    subtitle = "July 2026"
)

p2 <- plot_rgb(
    post_comp,
    title = "Post-fire RGB",
    subtitle = "August 2026"
)

p3 <- plot_dnbr(
    dnbr,
    title = "Difference NBR"
)

p4 <- plot_severity(
    severity,
    title = "Burn severity"
)

p1 <- p1 +
    ggplot2::theme(
        legend.position = "none"
    )

p2 <- p2 +
    ggplot2::theme(
        legend.position = "none"
    )

```


```{r}



figure <- patchwork::wrap_plots(
    p1,
    p2,
    p3,
    p4,
    ncol = 2,
    guides = "collect"
) +
    
    patchwork::plot_layout(
        guides = "collect"
    ) +

    patchwork::plot_annotation(

        title = "Wildfire burn assessment",

        subtitle = "Sentinel-2 composite imagery and burn severity",

        theme = ggplot2::theme(

            plot.title = ggplot2::element_text(
                face = "bold",
                size = 18
            ),

            plot.subtitle = ggplot2::element_text(
                size = 12
            )

        )

    )



```

```{r, eval=FALSE}

figure

ggplot2::ggsave(
    "man/figures/workflow.png",
    figure,
    width = 10,
    height = 8,
    dpi = 200
)

ggplot2::ggsave(
    "man/figures/workflow.pdf",
    figure,
    width = 10,
    height = 8
)

```

