---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

library(sentinelBurnR)
library(terra)

cache_dir <- file.path(
    path.expand("~"),
    "Library",
    "CloudStorage",
    "GoogleDrive-julian.flowers12@gmail.com",
    "My Drive",
    "Sentinel2",
    "cache"
)

```

# Introduction

`sentinelBurnR` provides an end-to-end workflow for producing burn severity maps from Sentinel-2 imagery.

This vignette demonstrates the complete workflow:

1. Create and read an Area of Interest (AOI)
2. Search the Sentinel-2 catalogue
3. Download imagery
4. Build cloud-free composites
5. Calculate NBR
6. Calculate dNBR

## Create an AOI

```{r}

lat <- 52.2522
lon <- 1.6271
  
buffer <- 1000

```


```{r}
aoi <- create_aoi(lat = lat, lon = lon, radius = buffer)
```


## Read an AOI

```{r}


aoi <- read_aoi(aoi$geometry)

```

## Search Sentinel-2

```{r}

pre_images <- search_s2(
    aoi,
    start = "2026-05-01",
    end = "2026-06-15"
)

post_images <- search_s2(
    aoi,
    start = "2026-07-30",
    end = "2026-08-09"
)


```

## Download

```{r, eval = TRUE}

## Download imagery

pre_collection <- download_s2(

    pre_images,
    
    limit = 2,

    workers = 6

)

post_collection <- download_s2(

    post_images,
    
    limit = 2,

    workers = 6

)
```

## Composite

```{r, eval=TRUE}

pre_comp <- build_composite(
    pre_collection
)

post_comp <- build_composite(
    post_collection
)

```

## NBR

```{r}

pre_nbr <- calc_nbr(
    pre_comp
)

post_nbr <- calc_nbr(
    post_comp
)

```

## dNBR

```{r}

dnbr <- calc_dnbr(

    pre_nbr,

    post_nbr

)

plot(dnbr)


```


